---
title: "CitiBike"
author: "Tanisha Pardasani"
date: "12/13/2020"
output: html_document
---
```{r}
library(tidyverse)
library(dplyr) 
library(ggplot2)
library(lubridate)
```


```{r}
citidata <- read.csv("fulldata.csv")
```

```{r}
head(citidata)
```



What are the demographics of riders?
```{r}
#convert genders numbers to names
citidata$gender <- as.factor(citidata$gender)

citidata2 <-
    citidata %>%
    mutate (gender1 = ifelse(citidata$gender==1,"Male",ifelse(citidata$gender==2,"Female","Unknown")) ) 
    
#table of average trip duration and average age by gender
demographics <- 
    citidata2 %>%
    group_by (gender1) %>%
    mutate (age = 2019 - birth.year) %>%
    summarize (count = n(), average_trip_duration = mean (tripduration),
                average_age = mean(age) ) %>%
    arrange(desc(count))
demographics
    
```
The age of riders for each gender are fairly similar. On average, females have longer trip durations of about 165 seconds.



```{r}
ggplot (demographics) + 
    geom_line(mapping = aes(x=gender1, y=count), color="green") +
    geom_line(mapping = aes(x=gender1, y=average_trip_duration), color="blue") +
    geom_line(mapping = aes(x=gender1, y=average_age), color="red")  +
    ggtitle( "Data by Gender") +
    xlab (" Gender") +
    ylab ("Count (Green), Average Trip Duration (Blue), Average Age (Red)")
```

What station is the start of most trips?
```{r}
start_station <- 
  citidata %>% 
  group_by(start.station.name) %>% 
  summarize(bikes_taken=n()) %>% 
  arrange(desc(bikes_taken)) 
head (start_station)
```

What station is the end of most trips?
```{r}
end_station <- 
  citidata %>% 
  group_by(end.station.name) %>% 
  summarize(bikes_returned=n()) %>% 
  arrange(desc(bikes_returned)) 
head (end_station)
```
Pershing Square North and E 17 St & Broadway are the most popular start and end stations. 


Which stations have more bikes leave than arrive?
```{r}
merged <- merge ( start_station, end_station, by.x = "start.station.name", by.y = "end.station.name")

new_table <-
  merged %>%
  filter ( bikes_taken > bikes_returned) %>%
  summarize ( "Station Name"=start.station.name , "Bikes Taken" = bikes_taken , "Bikes Returned" =     
              bikes_returned )

head(new_table)
```

About half (458 out of 927 total stations) of the stations have more bikes taken than returned. 
